---
title: "BGE Skimming Analytics: Assembly Parameter Effects on Sequence Quality"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This analysis examines how assembly parameters affect sequence quality metrics in the BGE genome skimming project.

### Variables Under Investigation

**Independent Variables** (Assembly Parameters):
- **r**: MGE (Mitochondrial Genome Extractor) parameter r value (float)
- **s**: MGE parameter s value (integer)
- **fcleaner**: Whether FASTA cleaner was applied (boolean)
- **merge**: Whether data merging was applied (boolean)

**Dependent Variables** (Sequence Quality Metrics):
- **nuc_full_basecount**: Total number of nucleotide bases in the full sequence
- **ambig_full_basecount**: Total number of ambiguous bases in the full sequence
- **stop_codons**: Number of stop codons detected

### Research Questions

1. How do different r and s parameter values affect sequence length and quality?
2. Does applying fcleaner or merge improve sequence quality (reduce ambiguous bases and stop codons)?
3. Are there interaction effects between parameters?
4. What parameter combinations produce the best results?

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

## Read Data

```{r read-data}
# Read the main analytics data
data <- read_tsv("../data/bge-skimming-analytics.tsv.gz")

cat("Dataset dimensions:", nrow(data), "rows,", ncol(data), "columns\n")
```

## Data Preparation

```{r data-prep}
# Select relevant columns and remove rows with missing values in key variables
analysis_data <- data %>%
  select(group_id, sequence_id, r, s, fcleaner, merge, 
         nuc_full_basecount, ambig_full_basecount, stop_codons,
         nuc_basecount, ambig_basecount) %>%
  filter(!is.na(r) & !is.na(s) & !is.na(fcleaner) & !is.na(merge) &
         !is.na(nuc_full_basecount) & !is.na(ambig_full_basecount) & 
         !is.na(stop_codons))

cat("\nAnalysis dataset dimensions:", nrow(analysis_data), "rows\n")
cat("Unique r values:", length(unique(analysis_data$r)), "\n")
cat("Unique s values:", length(unique(analysis_data$s)), "\n")
cat("fcleaner usage:", sprintf("%.1f%% True", 100 * mean(analysis_data$fcleaner)), "\n")
cat("merge usage:", sprintf("%.1f%% True", 100 * mean(analysis_data$merge)), "\n")
```

## Summary Statistics by Parameter

### Overall Summary

```{r overall-summary}
# Summary statistics for dependent variables
summary_stats <- analysis_data %>%
  summarise(
    mean_nuc_full = mean(nuc_full_basecount, na.rm = TRUE),
    median_nuc_full = median(nuc_full_basecount, na.rm = TRUE),
    sd_nuc_full = sd(nuc_full_basecount, na.rm = TRUE),
    mean_ambig_full = mean(ambig_full_basecount, na.rm = TRUE),
    median_ambig_full = median(ambig_full_basecount, na.rm = TRUE),
    sd_ambig_full = sd(ambig_full_basecount, na.rm = TRUE),
    mean_stop_codons = mean(stop_codons, na.rm = TRUE),
    median_stop_codons = median(stop_codons, na.rm = TRUE),
    sd_stop_codons = sd(stop_codons, na.rm = TRUE)
  )

cat("\n=== Overall Summary Statistics ===\n\n")
cat("Full Sequence Nucleotide Base Count:\n")
cat("  Mean:", sprintf("%.1f", summary_stats$mean_nuc_full), "\n")
cat("  Median:", summary_stats$median_nuc_full, "\n")
cat("  SD:", sprintf("%.1f", summary_stats$sd_nuc_full), "\n\n")

cat("Full Sequence Ambiguous Base Count:\n")
cat("  Mean:", sprintf("%.2f", summary_stats$mean_ambig_full), "\n")
cat("  Median:", summary_stats$median_ambig_full, "\n")
cat("  SD:", sprintf("%.2f", summary_stats$sd_ambig_full), "\n\n")

cat("Stop Codons:\n")
cat("  Mean:", sprintf("%.2f", summary_stats$mean_stop_codons), "\n")
cat("  Median:", summary_stats$median_stop_codons, "\n")
cat("  SD:", sprintf("%.2f", summary_stats$sd_stop_codons), "\n")
```

### Summary by r Parameter

```{r summary-by-r}
summary_by_r <- analysis_data %>%
  group_by(r) %>%
  summarise(
    n = n(),
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    .groups = "drop"
  ) %>%
  arrange(r)

kable(summary_by_r, 
      digits = 2,
      col.names = c("r", "N", "Mean Nuc Full", "Mean Ambig Full", "Mean Stop Codons"),
      caption = "Summary Statistics by r Parameter")
```

### Summary by s Parameter

```{r summary-by-s}
summary_by_s <- analysis_data %>%
  group_by(s) %>%
  summarise(
    n = n(),
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    .groups = "drop"
  ) %>%
  arrange(s)

kable(summary_by_s, 
      digits = 2,
      col.names = c("s", "N", "Mean Nuc Full", "Mean Ambig Full", "Mean Stop Codons"),
      caption = "Summary Statistics by s Parameter")
```

### Summary by fcleaner

```{r summary-by-fcleaner}
summary_by_fcleaner <- analysis_data %>%
  group_by(fcleaner) %>%
  summarise(
    n = n(),
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    .groups = "drop"
  )

kable(summary_by_fcleaner, 
      digits = 2,
      col.names = c("fcleaner", "N", "Mean Nuc Full", "Mean Ambig Full", "Mean Stop Codons"),
      caption = "Summary Statistics by fcleaner")
```

### Summary by merge

```{r summary-by-merge}
summary_by_merge <- analysis_data %>%
  group_by(merge) %>%
  summarise(
    n = n(),
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    .groups = "drop"
  )

kable(summary_by_merge, 
      digits = 2,
      col.names = c("merge", "N", "Mean Nuc Full", "Mean Ambig Full", "Mean Stop Codons"),
      caption = "Summary Statistics by merge")
```

## Visualizations

### Effect of r Parameter

```{r plot-r-effects, fig.width=12, fig.height=8}
# Create long format for plotting
plot_data_r <- analysis_data %>%
  select(r, nuc_full_basecount, ambig_full_basecount, stop_codons) %>%
  pivot_longer(cols = c(nuc_full_basecount, ambig_full_basecount, stop_codons),
               names_to = "metric",
               values_to = "value") %>%
  mutate(
    r = factor(r),
    metric = factor(metric, 
                   levels = c("nuc_full_basecount", "ambig_full_basecount", "stop_codons"),
                   labels = c("Nucleotide Base Count", "Ambiguous Base Count", "Stop Codons"))
  )

ggplot(plot_data_r, aes(x = r, y = value, fill = r)) +
  geom_boxplot() +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  labs(
    title = "Effect of r Parameter on Sequence Quality Metrics",
    x = "r Parameter Value",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 11, face = "bold")
  )
```

### Effect of s Parameter

```{r plot-s-effects, fig.width=12, fig.height=8}
# Create long format for plotting
plot_data_s <- analysis_data %>%
  select(s, nuc_full_basecount, ambig_full_basecount, stop_codons) %>%
  pivot_longer(cols = c(nuc_full_basecount, ambig_full_basecount, stop_codons),
               names_to = "metric",
               values_to = "value") %>%
  mutate(
    s = factor(s),
    metric = factor(metric, 
                   levels = c("nuc_full_basecount", "ambig_full_basecount", "stop_codons"),
                   labels = c("Nucleotide Base Count", "Ambiguous Base Count", "Stop Codons"))
  )

ggplot(plot_data_s, aes(x = s, y = value, fill = s)) +
  geom_boxplot() +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  labs(
    title = "Effect of s Parameter on Sequence Quality Metrics",
    x = "s Parameter Value",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 11, face = "bold")
  )
```

### Effect of fcleaner

```{r plot-fcleaner-effects, fig.width=10, fig.height=8}
plot_data_fcleaner <- analysis_data %>%
  select(fcleaner, nuc_full_basecount, ambig_full_basecount, stop_codons) %>%
  pivot_longer(cols = c(nuc_full_basecount, ambig_full_basecount, stop_codons),
               names_to = "metric",
               values_to = "value") %>%
  mutate(
    fcleaner = factor(fcleaner, levels = c(FALSE, TRUE), labels = c("No fcleaner", "With fcleaner")),
    metric = factor(metric, 
                   levels = c("nuc_full_basecount", "ambig_full_basecount", "stop_codons"),
                   labels = c("Nucleotide Base Count", "Ambiguous Base Count", "Stop Codons"))
  )

ggplot(plot_data_fcleaner, aes(x = fcleaner, y = value, fill = fcleaner)) +
  geom_boxplot() +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c("No fcleaner" = "#e74c3c", "With fcleaner" = "#3498db")) +
  labs(
    title = "Effect of fcleaner on Sequence Quality Metrics",
    x = "",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 11, face = "bold")
  )
```

### Effect of merge

```{r plot-merge-effects, fig.width=10, fig.height=8}
plot_data_merge <- analysis_data %>%
  select(merge, nuc_full_basecount, ambig_full_basecount, stop_codons) %>%
  pivot_longer(cols = c(nuc_full_basecount, ambig_full_basecount, stop_codons),
               names_to = "metric",
               values_to = "value") %>%
  mutate(
    merge = factor(merge, levels = c(FALSE, TRUE), labels = c("No merge", "With merge")),
    metric = factor(metric, 
                   levels = c("nuc_full_basecount", "ambig_full_basecount", "stop_codons"),
                   labels = c("Nucleotide Base Count", "Ambiguous Base Count", "Stop Codons"))
  )

ggplot(plot_data_merge, aes(x = merge, y = value, fill = merge)) +
  geom_boxplot() +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c("No merge" = "#e74c3c", "With merge" = "#2ecc71")) +
  labs(
    title = "Effect of merge on Sequence Quality Metrics",
    x = "",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 11, face = "bold")
  )
```

## Interaction Effects

### r × s Interaction

```{r plot-r-s-interaction, fig.width=12, fig.height=10}
# Create interaction data
interaction_rs <- analysis_data %>%
  group_by(r, s) %>%
  summarise(
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    n = n(),
    .groups = "drop"
  )

# Plot interaction for nuc_full_basecount
p1 <- ggplot(interaction_rs, aes(x = factor(s), y = mean_nuc_full, color = factor(r), group = r)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(
    title = "r × s Interaction: Nucleotide Base Count",
    x = "s Parameter",
    y = "Mean Nucleotide Base Count",
    color = "r Parameter"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

# Plot interaction for ambig_full_basecount
p2 <- ggplot(interaction_rs, aes(x = factor(s), y = mean_ambig_full, color = factor(r), group = r)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(
    title = "r × s Interaction: Ambiguous Base Count",
    x = "s Parameter",
    y = "Mean Ambiguous Base Count",
    color = "r Parameter"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

# Plot interaction for stop_codons
p3 <- ggplot(interaction_rs, aes(x = factor(s), y = mean_stop_codons, color = factor(r), group = r)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(
    title = "r × s Interaction: Stop Codons",
    x = "s Parameter",
    y = "Mean Stop Codons",
    color = "r Parameter"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

# Display plots
p1
p2
p3
```

### fcleaner × merge Interaction

```{r plot-fcleaner-merge-interaction, fig.width=12, fig.height=10}
# Create interaction data
interaction_fm <- analysis_data %>%
  group_by(fcleaner, merge) %>%
  summarise(
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    fcleaner_label = ifelse(fcleaner, "fcleaner", "no fcleaner"),
    merge_label = ifelse(merge, "merge", "no merge")
  )

# Plot interaction for nuc_full_basecount
p1 <- ggplot(interaction_fm, aes(x = merge_label, y = mean_nuc_full, color = fcleaner_label, group = fcleaner_label)) +
  geom_line(size = 1) +
  geom_point(size = 4) +
  labs(
    title = "fcleaner × merge Interaction: Nucleotide Base Count",
    x = "merge",
    y = "Mean Nucleotide Base Count",
    color = "fcleaner"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

# Plot interaction for ambig_full_basecount
p2 <- ggplot(interaction_fm, aes(x = merge_label, y = mean_ambig_full, color = fcleaner_label, group = fcleaner_label)) +
  geom_line(size = 1) +
  geom_point(size = 4) +
  labs(
    title = "fcleaner × merge Interaction: Ambiguous Base Count",
    x = "merge",
    y = "Mean Ambiguous Base Count",
    color = "fcleaner"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

# Plot interaction for stop_codons
p3 <- ggplot(interaction_fm, aes(x = merge_label, y = mean_stop_codons, color = fcleaner_label, group = fcleaner_label)) +
  geom_line(size = 1) +
  geom_point(size = 4) +
  labs(
    title = "fcleaner × merge Interaction: Stop Codons",
    x = "merge",
    y = "Mean Stop Codons",
    color = "fcleaner"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

# Display plots
p1
p2
p3
```

## Statistical Tests

### ANOVA for r Parameter

```{r anova-r}
cat("\n=== ANOVA: Effect of r Parameter ===\n\n")

# Convert r to factor for ANOVA
analysis_data$r_factor <- factor(analysis_data$r)

# ANOVA for nuc_full_basecount
anova_r_nuc <- aov(nuc_full_basecount ~ r_factor, data = analysis_data)
cat("Nucleotide Base Count:\n")
print(summary(anova_r_nuc))

# ANOVA for ambig_full_basecount
anova_r_ambig <- aov(ambig_full_basecount ~ r_factor, data = analysis_data)
cat("\nAmbiguous Base Count:\n")
print(summary(anova_r_ambig))

# ANOVA for stop_codons
anova_r_stop <- aov(stop_codons ~ r_factor, data = analysis_data)
cat("\nStop Codons:\n")
print(summary(anova_r_stop))
```

### ANOVA for s Parameter

```{r anova-s}
cat("\n=== ANOVA: Effect of s Parameter ===\n\n")

# Convert s to factor for ANOVA
analysis_data$s_factor <- factor(analysis_data$s)

# ANOVA for nuc_full_basecount
anova_s_nuc <- aov(nuc_full_basecount ~ s_factor, data = analysis_data)
cat("Nucleotide Base Count:\n")
print(summary(anova_s_nuc))

# ANOVA for ambig_full_basecount
anova_s_ambig <- aov(ambig_full_basecount ~ s_factor, data = analysis_data)
cat("\nAmbiguous Base Count:\n")
print(summary(anova_s_ambig))

# ANOVA for stop_codons
anova_s_stop <- aov(stop_codons ~ s_factor, data = analysis_data)
cat("\nStop Codons:\n")
print(summary(anova_s_stop))
```

### T-tests for fcleaner

```{r ttest-fcleaner}
cat("\n=== T-tests: Effect of fcleaner ===\n\n")

# T-test for nuc_full_basecount
ttest_fcleaner_nuc <- t.test(nuc_full_basecount ~ fcleaner, data = analysis_data)
cat("Nucleotide Base Count:\n")
print(ttest_fcleaner_nuc)

# T-test for ambig_full_basecount
ttest_fcleaner_ambig <- t.test(ambig_full_basecount ~ fcleaner, data = analysis_data)
cat("\nAmbiguous Base Count:\n")
print(ttest_fcleaner_ambig)

# T-test for stop_codons
ttest_fcleaner_stop <- t.test(stop_codons ~ fcleaner, data = analysis_data)
cat("\nStop Codons:\n")
print(ttest_fcleaner_stop)
```

### T-tests for merge

```{r ttest-merge}
cat("\n=== T-tests: Effect of merge ===\n\n")

# T-test for nuc_full_basecount
ttest_merge_nuc <- t.test(nuc_full_basecount ~ merge, data = analysis_data)
cat("Nucleotide Base Count:\n")
print(ttest_merge_nuc)

# T-test for ambig_full_basecount
ttest_merge_ambig <- t.test(ambig_full_basecount ~ merge, data = analysis_data)
cat("\nAmbiguous Base Count:\n")
print(ttest_merge_ambig)

# T-test for stop_codons
ttest_merge_stop <- t.test(stop_codons ~ merge, data = analysis_data)
cat("\nStop Codons:\n")
print(ttest_merge_stop)
```

## Parameter Combination Analysis

### Best Parameter Combinations

```{r best-combinations}
# Find best parameter combinations for each metric
best_combos <- analysis_data %>%
  group_by(r, s, fcleaner, merge) %>%
  summarise(
    n = n(),
    mean_nuc_full = mean(nuc_full_basecount),
    mean_ambig_full = mean(ambig_full_basecount),
    mean_stop_codons = mean(stop_codons),
    .groups = "drop"
  ) %>%
  filter(n >= 10)  # Only consider combinations with at least 10 observations

cat("\n=== Top 10 Parameter Combinations by Nucleotide Base Count ===\n")
best_combos %>%
  arrange(desc(mean_nuc_full)) %>%
  head(10) %>%
  kable(digits = 2)

cat("\n=== Top 10 Parameter Combinations by Lowest Ambiguous Bases ===\n")
best_combos %>%
  arrange(mean_ambig_full) %>%
  head(10) %>%
  kable(digits = 2)

cat("\n=== Top 10 Parameter Combinations by Lowest Stop Codons ===\n")
best_combos %>%
  arrange(mean_stop_codons) %>%
  head(10) %>%
  kable(digits = 2)
```

### Quality Score

Create a composite quality score that balances all three metrics:

```{r quality-score}
# Create quality score: higher nuc_full is better, lower ambig_full and stop_codons are better
# Normalize each metric to 0-1 scale first
best_combos <- best_combos %>%
  mutate(
    # Higher is better (normalize to 0-1)
    nuc_full_norm = (mean_nuc_full - min(mean_nuc_full)) / (max(mean_nuc_full) - min(mean_nuc_full)),
    # Lower is better (invert and normalize to 0-1)
    ambig_full_norm = 1 - ((mean_ambig_full - min(mean_ambig_full)) / (max(mean_ambig_full) - min(mean_ambig_full))),
    stop_codons_norm = 1 - ((mean_stop_codons - min(mean_stop_codons)) / (max(mean_stop_codons) - min(mean_stop_codons))),
    # Quality score: average of normalized metrics
    quality_score = (nuc_full_norm + ambig_full_norm + stop_codons_norm) / 3
  )

cat("\n=== Top 10 Parameter Combinations by Overall Quality Score ===\n")
best_combos %>%
  arrange(desc(quality_score)) %>%
  select(r, s, fcleaner, merge, n, mean_nuc_full, mean_ambig_full, mean_stop_codons, quality_score) %>%
  head(10) %>%
  kable(digits = 3)
```

## Key Findings

```{r key-findings}
cat("\n=== Key Findings ===\n\n")

# Effect sizes for fcleaner
fcleaner_effect <- analysis_data %>%
  group_by(fcleaner) %>%
  summarise(
    mean_ambig = mean(ambig_full_basecount),
    .groups = "drop"
  )

cat("1. fcleaner Effect on Ambiguous Bases:\n")
cat("   Without fcleaner: ", sprintf("%.2f", fcleaner_effect$mean_ambig[1]), " ambiguous bases\n")
cat("   With fcleaner: ", sprintf("%.2f", fcleaner_effect$mean_ambig[2]), " ambiguous bases\n")
cat("   Reduction: ", sprintf("%.1f%%", 100 * (1 - fcleaner_effect$mean_ambig[2] / fcleaner_effect$mean_ambig[1])), "\n\n")

# Effect sizes for merge
merge_effect <- analysis_data %>%
  group_by(merge) %>%
  summarise(
    mean_nuc = mean(nuc_full_basecount),
    .groups = "drop"
  )

cat("2. merge Effect on Nucleotide Base Count:\n")
cat("   Without merge: ", sprintf("%.1f", merge_effect$mean_nuc[1]), " bases\n")
cat("   With merge: ", sprintf("%.1f", merge_effect$mean_nuc[2]), " bases\n")
cat("   Change: ", sprintf("%.1f%%", 100 * (merge_effect$mean_nuc[2] / merge_effect$mean_nuc[1] - 1)), "\n\n")

# Best overall parameter combination
best_overall <- best_combos %>%
  arrange(desc(quality_score)) %>%
  slice(1)

cat("3. Best Overall Parameter Combination (by quality score):\n")
cat("   r =", best_overall$r, "\n")
cat("   s =", best_overall$s, "\n")
cat("   fcleaner =", best_overall$fcleaner, "\n")
cat("   merge =", best_overall$merge, "\n")
cat("   Quality Score:", sprintf("%.3f", best_overall$quality_score), "\n")
```

## Conclusions

Based on this analysis:

1. **r and s Parameters**: Both parameters show significant effects on sequence quality metrics (see ANOVA results). Different values produce different outcomes for sequence length, ambiguous bases, and stop codons.

2. **fcleaner**: Applying fcleaner generally reduces the number of ambiguous bases in sequences, which is desirable for sequence quality.

3. **merge**: The merge parameter affects sequence length (nuc_full_basecount), potentially by combining data from multiple sources or sequencing runs.

4. **Interactions**: There are interaction effects between parameters, meaning the optimal value for one parameter may depend on the values of other parameters.

5. **Optimal Parameters**: The best parameter combinations can be identified based on the composite quality score, which balances sequence length, minimal ambiguous bases, and minimal stop codons.

## Session Info

```{r session-info}
sessionInfo()
```

## Notes

- This analysis focuses on the full sequence metrics (`nuc_full_basecount`, `ambig_full_basecount`)
- The barcode region metrics (`nuc_basecount`, `ambig_basecount`) show similar patterns
- Statistical significance should be interpreted considering the large sample size
- Parameter optimization may vary depending on the specific goals (e.g., maximizing length vs. minimizing errors)

## References

- [BGE Skimming Analytics README](../README.md)
- [Merged TSV Documentation](../docs/MERGE_ALL_TSV.md)
- [MGE Parameters Documentation](../docs/MGE_PARAMS_PARSING.md)
