---
title: "BGE Skimming Analytics with BOLD Metadata"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This document demonstrates how to read the compressed, frictionless BGE skimming analytics data and join it with BOLD metadata files to enrich the dataset with:

- **Sample ID**: Links validation results to BOLD metadata (from `lab.tsv`)
- **Collection Date**: When specimens were collected (from `collection_data.tsv`)
- **Institution Storing**: Source of specimens (from `voucher.tsv`)
- **Taxonomic Lineage**: Phylum through Subspecies levels (from `taxonomy.tsv`)

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Read Data Files

### Main Analytics Data

Read the compressed, frictionless data table directly from the gzip file:

```{r read-main-data}
# Read the main analytics data
analytics_data <- read_tsv("../data/bge-skimming-analytics.tsv.gz")

cat("Analytics data dimensions:", nrow(analytics_data), "rows,", 
    ncol(analytics_data), "columns\n")
```

### BOLD Metadata Files

According to `metadata/bold/README.md`, we need to join the following files:

1. **lab.tsv**: Join on `Process ID` to get `Sample ID`
2. **collection_data.tsv**: Join on `Sample ID` to get collection date
3. **voucher.tsv**: Join on `Sample ID` to get institution storing
4. **taxonomy.tsv**: Join on `Sample ID` to get taxonomic lineage

```{r read-bold-metadata}
# Read BOLD metadata files
lab_data <- read_tsv("../metadata/bold/lab.tsv")
collection_data <- read_tsv("../metadata/bold/collection_data.tsv")
voucher_data <- read_tsv("../metadata/bold/voucher.tsv")
taxonomy_data <- read_tsv("../metadata/bold/taxonomy.tsv")

cat("Lab data:", nrow(lab_data), "rows\n")
cat("Collection data:", nrow(collection_data), "rows\n")
cat("Voucher data:", nrow(voucher_data), "rows\n")
cat("Taxonomy data:", nrow(taxonomy_data), "rows\n")
```

## Join Data

### Step 1: Join with Lab Data

First, join with `lab.tsv` to link the `group_id` (which represents the Process ID in the analytics data) with the `Sample ID`:

```{r join-lab}
# Join analytics data with lab data
# group_id in analytics_data corresponds to "Process ID" in lab.tsv
enriched_data <- analytics_data %>%
  left_join(
    lab_data %>% select(`Process ID`, `Sample ID`, `Field ID`, BIN),
    by = c("group_id" = "Process ID")
  )

cat("After joining with lab data:", nrow(enriched_data), "rows,", 
    ncol(enriched_data), "columns\n")
cat("Rows with Sample ID:", sum(!is.na(enriched_data$`Sample ID`)), "\n")
```

### Step 2: Join with Collection Data

Join with `collection_data.tsv` to add collection date information:

```{r join-collection}
# Join with collection data to get collection date
enriched_data <- enriched_data %>%
  left_join(
    collection_data %>% 
      select(`Sample ID`, Collectors, `Collection Date`, `Country/Ocean`, `State/Province`),
    by = "Sample ID"
  )

cat("After joining with collection data:", ncol(enriched_data), "columns\n")
cat("Rows with Collection Date:", sum(!is.na(enriched_data$`Collection Date`)), "\n")
```

### Step 3: Join with Voucher Data

Join with `voucher.tsv` to add institution storing information:

```{r join-voucher}
# Join with voucher data to get institution storing
enriched_data <- enriched_data %>%
  left_join(
    voucher_data %>% 
      select(`Sample ID`, `Institution Storing`, `Museum ID`, `Collection Code`),
    by = "Sample ID",
    suffix = c("", ".voucher")
  )

cat("After joining with voucher data:", ncol(enriched_data), "columns\n")
cat("Rows with Institution Storing:", sum(!is.na(enriched_data$`Institution Storing`)), "\n")
```

### Step 4: Join with Taxonomy Data

Finally, join with `taxonomy.tsv` to add taxonomic lineage:

```{r join-taxonomy}
# Join with taxonomy data to get taxonomic lineage
enriched_data <- enriched_data %>%
  left_join(
    taxonomy_data %>% 
      select(`Sample ID`, Phylum, Class, Order, Family, Subfamily, Tribe, Genus, Species, Subspecies),
    by = "Sample ID"
  )

cat("Final enriched data:", nrow(enriched_data), "rows,", 
    ncol(enriched_data), "columns\n")
cat("Rows with Phylum:", sum(!is.na(enriched_data$Phylum)), "\n")
```

## Summary of Enriched Data

```{r summary}
# Display summary of key columns
cat("\n=== Summary of Enriched Data ===\n\n")

# Original analytics columns
cat("Original analytics data columns:", ncol(analytics_data), "\n")

# New columns added from BOLD metadata
new_columns <- setdiff(names(enriched_data), names(analytics_data))
cat("New columns added from BOLD metadata:", length(new_columns), "\n")
cat("\nNew columns:\n")
print(new_columns)

# Coverage statistics
cat("\n=== Data Coverage ===\n")
cat("Rows with Sample ID:", 
    sprintf("%.1f%%", 100 * sum(!is.na(enriched_data$`Sample ID`)) / nrow(enriched_data)), "\n")
cat("Rows with Collection Date:", 
    sprintf("%.1f%%", 100 * sum(!is.na(enriched_data$`Collection Date`)) / nrow(enriched_data)), "\n")
cat("Rows with Institution Storing:", 
    sprintf("%.1f%%", 100 * sum(!is.na(enriched_data$`Institution Storing`)) / nrow(enriched_data)), "\n")
cat("Rows with Taxonomic Data:", 
    sprintf("%.1f%%", 100 * sum(!is.na(enriched_data$Phylum)) / nrow(enriched_data)), "\n")
```

## Data Preview

```{r preview}
# Show a sample of the enriched data with key columns
enriched_data %>%
  select(group_id, `Sample ID`, sequence_id, 
         `Collection Date`, `Institution Storing`,
         Phylum, Class, Order, Family, Genus, Species) %>%
  head(10) %>%
  kable(caption = "Sample of Enriched Data with BOLD Metadata")
```

## Example Visualizations

### Collection Dates Over Time

Visualize when specimens were collected (requires date parsing):

```{r collection-dates, fig.width=10, fig.height=6}
# Parse collection dates (format: dd-Mmm-yyyy)
enriched_data_with_dates <- enriched_data %>%
  filter(!is.na(`Collection Date`)) %>%
  mutate(
    Collection_Year = as.integer(sub(".*-(\\d{4})$", "\\1", `Collection Date`))
  ) %>%
  filter(!is.na(Collection_Year))

if (nrow(enriched_data_with_dates) > 0) {
  ggplot(enriched_data_with_dates, aes(x = Collection_Year)) +
    geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
    labs(
      title = "Distribution of Specimen Collection Years",
      x = "Collection Year",
      y = "Number of Specimens"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
} else {
  cat("No collection date data available for visualization\n")
}
```

### Taxonomic Distribution

Visualize the distribution of specimens across phyla:

```{r taxonomy-distribution, fig.width=10, fig.height=6}
if (sum(!is.na(enriched_data$Phylum)) > 0) {
  enriched_data %>%
    filter(!is.na(Phylum)) %>%
    count(Phylum) %>%
    arrange(desc(n)) %>%
    head(10) %>%
    ggplot(aes(x = reorder(Phylum, n), y = n)) +
    geom_col(fill = "forestgreen") +
    coord_flip() +
    labs(
      title = "Top 10 Phyla in Dataset",
      x = "Phylum",
      y = "Number of Specimens"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
} else {
  cat("No taxonomic data available for visualization\n")
}
```

### Institution Distribution

Visualize specimens by storing institution:

```{r institution-distribution, fig.width=10, fig.height=6}
if (sum(!is.na(enriched_data$`Institution Storing`)) > 0) {
  enriched_data %>%
    filter(!is.na(`Institution Storing`)) %>%
    count(`Institution Storing`) %>%
    arrange(desc(n)) %>%
    ggplot(aes(x = reorder(`Institution Storing`, n), y = n)) +
    geom_col(fill = "coral") +
    coord_flip() +
    labs(
      title = "Specimens by Storing Institution",
      x = "Institution",
      y = "Number of Specimens"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
} else {
  cat("No institution data available for visualization\n")
}
```

## Export Enriched Data

Optionally, export the enriched data for further analysis:

```{r export, eval=FALSE}
# Save enriched data as compressed TSV
write_tsv(enriched_data, "../data/bge-skimming-analytics-enriched.tsv.gz")

cat("Enriched data saved to: data/bge-skimming-analytics-enriched.tsv.gz\n")
```

## Session Info

```{r session-info}
sessionInfo()
```

## Notes

- The `group_id` column in the analytics data corresponds to the `Process ID` in BOLD metadata
- Left joins are used to preserve all rows from the analytics data
- Some rows may not have matching BOLD metadata (NAs in joined columns)
- Collection dates are in dd-Mmm-yyyy format (e.g., "23-Aug-1983")
- Lower taxonomic levels may be empty if specimens are not identified to that level

## References

- [BOLD Metadata README](../metadata/bold/README.md)
- [Merged TSV Documentation](../docs/MERGE_ALL_TSV.md)
- [Frictionless Data Specification](https://frictionlessdata.io/)
