---
title: "Collection Localities Map"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This analysis visualizes the geographic distribution of BGE specimen collection localities across Europe. The map shows GPS coordinates from the BOLD dataset (DSBGEUROPE), which contains both genome skimming and fresh sequencing barcoding efforts.

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
```

## Read Collection Data

```{r read-data}
# Read the collection_data.tsv file from BOLD metadata
collection_data <- read_tsv("../metadata/bold/collection_data.tsv", show_col_types = FALSE)

cat("Collection data dimensions:", nrow(collection_data), "rows,", 
    ncol(collection_data), "columns\n")
```

## Data Processing

### Filter for Valid GPS Coordinates

```{r process-data}
# Filter to rows with valid GPS coordinates (non-empty Lat and Lon)
collection_with_gps <- collection_data %>%
  filter(!is.na(Lat) & !is.na(Lon) & Lat != "" & Lon != "")

# Convert to numeric (in case they were read as character)
collection_with_gps <- collection_with_gps %>%
  mutate(
    Latitude = as.numeric(Lat),
    Longitude = as.numeric(Lon)
  ) %>%
  filter(!is.na(Latitude) & !is.na(Longitude))

cat("Records with valid GPS coordinates:", nrow(collection_with_gps), "\n")
cat("Latitude range:", min(collection_with_gps$Latitude), "to", 
    max(collection_with_gps$Latitude), "\n")
cat("Longitude range:", min(collection_with_gps$Longitude), "to", 
    max(collection_with_gps$Longitude), "\n")
```

## Map Visualization

### Create Map of Europe with Collection Localities

```{r map-visualization, fig.width=12, fig.height=10}
# Get European country borders
europe <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf")

# Create the map
ggplot() +
  # Add country borders
  geom_sf(data = europe, fill = "white", color = "gray50", linewidth = 0.3) +
  # Add collection points
  geom_point(data = collection_with_gps, 
             aes(x = Longitude, y = Latitude),
             color = "red", 
             size = 0.5, 
             alpha = 0.6) +
  # Set coordinate limits to focus on Europe
  coord_sf(xlim = c(-25, 50), ylim = c(30, 75), expand = FALSE) +
  # Add labels and theme
  labs(
    title = "BGE Specimen Collection Localities in Europe",
    subtitle = paste0("Showing ", format(nrow(collection_with_gps), big.mark = ","), 
                     " collection localities with GPS coordinates"),
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.2),
    panel.background = element_rect(fill = "lightblue", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

## Summary Statistics

### Geographic Coverage

```{r summary-stats}
# Calculate summary statistics
summary_stats <- collection_with_gps %>%
  summarise(
    Total_Localities = n(),
    Unique_Countries = n_distinct(`Country/Ocean`, na.rm = TRUE),
    Latitude_Min = min(Latitude),
    Latitude_Max = max(Latitude),
    Longitude_Min = min(Longitude),
    Longitude_Max = max(Longitude)
  )

kable(summary_stats, 
      caption = "Geographic Coverage Statistics",
      col.names = c("Total Localities", "Unique Countries", 
                    "Min Latitude", "Max Latitude", 
                    "Min Longitude", "Max Longitude"))
```

### Top Countries by Number of Collection Localities

```{r country-summary}
# Count localities by country
country_counts <- collection_with_gps %>%
  filter(!is.na(`Country/Ocean`) & `Country/Ocean` != "") %>%
  group_by(`Country/Ocean`) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  head(20)

kable(country_counts, 
      caption = "Top 20 Countries by Number of Collection Localities",
      col.names = c("Country/Ocean", "Number of Localities"))
```

## Conclusion

This map visualizes the extensive geographic coverage of the BGE barcoding project across Europe. The distribution shows where specimen collection efforts have been concentrated, providing insight into the spatial coverage of the dataset.
