---
title: "Successfully Sequenced Specimens by Institution"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This analysis examines COI-5P sequencing success across different institutions in the BGE project. For each institution, we calculate:

- **Registered specimens**: Specimens with any data in the `COI-5P Seq. Length` column (indicating they were registered for COI-5P sequencing)
- **Successfully sequenced**: Specimens where `COI-5P Seq. Length` is not `0[n]` (indicating a sequence of any length was recovered and uploaded to BOLD)

The analysis focuses on the number of successfully sequenced specimens, with institutions ordered by their successful output.

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

## Read Lab Data

```{r read-data}
# Read the lab.tsv file from BOLD metadata
lab_data <- read_tsv("../metadata/bold/lab.tsv", show_col_types = FALSE)

cat("Lab data dimensions:", nrow(lab_data), "rows,", 
    ncol(lab_data), "columns\n")
```

## Data Processing

### Group by Institution and Calculate Metrics

```{r process-data}
# Process the data to calculate registered and successfully sequenced specimens
institution_summary <- lab_data %>%
  # Filter to rows with Institution data
  filter(!is.na(Institution) & Institution != "" & Institution != "Institution") %>%
  # Group by institution
  group_by(Institution) %>%
  # Calculate metrics
  summarise(
    # Registered: any non-empty data in COI-5P Seq. Length
    registered = sum(!is.na(`COI-5P Seq. Length`) & `COI-5P Seq. Length` != ""),
    # Successfully sequenced: not "0[n]" and not empty
    successful = sum(!is.na(`COI-5P Seq. Length`) & 
                     `COI-5P Seq. Length` != "" & 
                     `COI-5P Seq. Length` != "0[n]"),
    .groups = "drop"
  ) %>%
  # Order by decreasing number of successfully sequenced specimens
  arrange(desc(successful))

# Display summary statistics
cat("\n=== Summary Statistics ===\n\n")
cat("Total institutions:", nrow(institution_summary), "\n")
cat("Total registered specimens:", sum(institution_summary$registered), "\n")
cat("Total successfully sequenced:", sum(institution_summary$successful), "\n")
```

## Data Preview

```{r preview}
# Show top 20 institutions
institution_summary %>%
  head(20) %>%
  kable(caption = "Top 20 Institutions by Number of Successfully Sequenced Specimens")
```

## Visualization

### Bar Chart of Successfully Sequenced Specimens by Institution

```{r sequencing-success-chart, fig.width=12, fig.height=10}
# Create bar chart showing only successfully sequenced specimens
ggplot(institution_summary, aes(x = reorder(Institution, successful), y = successful)) +
  geom_col(fill = "#2ecc71") +
  coord_flip() +
  labs(
    title = "Successfully Sequenced Specimens by Institution",
    subtitle = "Institutions ordered by number of successfully sequenced specimens",
    x = "Institution",
    y = "Number of Successfully Sequenced Specimens"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 8)
  )
```

### Top 15 Institutions by Successfully Sequenced Specimens

```{r top-institutions-chart, fig.width=12, fig.height=8}
# Focus on top 15 institutions for better readability
top_15_data <- institution_summary %>%
  head(15)

ggplot(top_15_data, aes(x = reorder(Institution, successful), y = successful)) +
  geom_col(fill = "#2ecc71") +
  geom_text(
    aes(label = successful),
    hjust = -0.1,
    size = 3
  ) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 15 Institutions: Successfully Sequenced Specimens",
    subtitle = "Numbers show count of successfully sequenced specimens",
    x = "Institution",
    y = "Number of Successfully Sequenced Specimens"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 9)
  )
```

## Detailed Statistics by Institution

### Statistics Table

```{r statistics-table}
# Show top 30 institutions by successfully sequenced specimens
institution_summary %>%
  head(30) %>%
  select(Institution, registered, successful) %>%
  kable(
    col.names = c("Institution", "Total Registered", "Successfully Sequenced"),
    caption = "Top 30 Institutions by Successfully Sequenced Specimens"
  )
```

### Distribution of Successfully Sequenced Specimens

```{r successful-distribution, fig.width=10, fig.height=6}
ggplot(institution_summary, aes(x = successful)) +
  geom_histogram(binwidth = 50, fill = "#2ecc71", color = "white") +
  labs(
    title = "Distribution of Successfully Sequenced Specimens Across Institutions",
    x = "Number of Successfully Sequenced Specimens",
    y = "Number of Institutions"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

## Session Info

```{r session-info}
sessionInfo()
```

## Notes

- **Registered specimens**: Specimens with any data in the `COI-5P Seq. Length` column
- **Successfully sequenced**: Specimens where `COI-5P Seq. Length` is not `0[n]` (meaning a sequence was recovered)
- The `0[n]` notation indicates 0 base pairs with n ambiguous bases (i.e., no sequence recovered)
- Institutions are ordered by the number of successfully sequenced specimens (descending)

