---
title: "COI-5P Sequencing Success by Institution"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This analysis examines COI-5P sequencing success rates across different institutions in the BGE project. For each institution, we calculate:

- **Registered specimens**: Specimens with any data in the `COI-5P Seq. Length` column (indicating they were registered for COI-5P sequencing)
- **Successfully sequenced**: Specimens where `COI-5P Seq. Length` is not `0[n]` (indicating a sequence of any length was recovered and uploaded to BOLD)

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

## Read Lab Data

```{r read-data}
# Read the lab.tsv file from BOLD metadata
lab_data <- read_tsv("../metadata/bold/lab.tsv", show_col_types = FALSE)

cat("Lab data dimensions:", nrow(lab_data), "rows,", 
    ncol(lab_data), "columns\n")
```

## Data Processing

### Group by Institution and Calculate Metrics

```{r process-data}
# Process the data to calculate registered and successfully sequenced specimens
institution_summary <- lab_data %>%
  # Filter to rows with Institution data
  filter(!is.na(Institution) & Institution != "" & Institution != "Institution") %>%
  # Group by institution
  group_by(Institution) %>%
  # Calculate metrics
  summarise(
    # Registered: any non-empty data in COI-5P Seq. Length
    registered = sum(!is.na(`COI-5P Seq. Length`) & `COI-5P Seq. Length` != ""),
    # Successfully sequenced: not "0[n]" and not empty
    successful = sum(!is.na(`COI-5P Seq. Length`) & 
                     `COI-5P Seq. Length` != "" & 
                     `COI-5P Seq. Length` != "0[n]"),
    .groups = "drop"
  ) %>%
  # Calculate failed (registered but not successful)
  mutate(
    failed = registered - successful
  ) %>%
  # Order by decreasing number of registered specimens
  arrange(desc(registered))

# Display summary statistics
cat("\n=== Summary Statistics ===\n\n")
cat("Total institutions:", nrow(institution_summary), "\n")
cat("Total registered specimens:", sum(institution_summary$registered), "\n")
cat("Total successfully sequenced:", sum(institution_summary$successful), "\n")
cat("Overall success rate:", 
    sprintf("%.1f%%", 100 * sum(institution_summary$successful) / sum(institution_summary$registered)), "\n")
```

## Data Preview

```{r preview}
# Show top 20 institutions
institution_summary %>%
  head(20) %>%
  kable(caption = "Top 20 Institutions by Number of Registered Specimens")
```

## Visualization

### Stacked Bar Chart of Sequencing Success by Institution

```{r sequencing-success-chart, fig.width=12, fig.height=10}
# Prepare data for stacked bar chart
chart_data <- institution_summary %>%
  # Convert to long format for stacking
  pivot_longer(
    cols = c(successful, failed),
    names_to = "status",
    values_to = "count"
  ) %>%
  # Create factor with desired order for stacking
  mutate(
    status = factor(status, 
                   levels = c("successful", "failed"),
                   labels = c("Successfully Sequenced", "Failed/Not Sequenced"))
  )

# Create the stacked bar chart
ggplot(chart_data, aes(x = reorder(Institution, registered), y = count, fill = status)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("Successfully Sequenced" = "#2ecc71", "Failed/Not Sequenced" = "#e74c3c"),
    name = "Status"
  ) +
  labs(
    title = "COI-5P Sequencing Success by Institution",
    subtitle = "Institutions ordered by total number of registered specimens",
    x = "Institution",
    y = "Number of Specimens"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    axis.text.y = element_text(size = 8)
  )
```

### Top 15 Institutions with Success Rates

```{r top-institutions-chart, fig.width=12, fig.height=8}
# Focus on top 15 institutions for better readability
top_15_data <- institution_summary %>%
  head(15) %>%
  pivot_longer(
    cols = c(successful, failed),
    names_to = "status",
    values_to = "count"
  ) %>%
  mutate(
    status = factor(status, 
                   levels = c("successful", "failed"),
                   labels = c("Successfully Sequenced", "Failed/Not Sequenced"))
  )

# Calculate success rate for annotation
top_15_rates <- institution_summary %>%
  head(15) %>%
  mutate(
    success_rate = 100 * successful / registered,
    label_text = sprintf("%.1f%%", success_rate)
  )

ggplot(top_15_data, aes(x = reorder(Institution, registered), y = count, fill = status)) +
  geom_col() +
  geom_text(
    data = top_15_rates,
    aes(x = Institution, y = registered, label = label_text, fill = NULL),
    hjust = -0.1,
    size = 3
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Successfully Sequenced" = "#2ecc71", "Failed/Not Sequenced" = "#e74c3c"),
    name = "Status"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 15 Institutions: COI-5P Sequencing Success",
    subtitle = "Numbers show success rate percentage",
    x = "Institution",
    y = "Number of Specimens"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    axis.text.y = element_text(size = 9)
  )
```

## Detailed Statistics by Institution

### Success Rates Table

```{r success-rates-table}
# Create a detailed table with success rates
detailed_stats <- institution_summary %>%
  mutate(
    success_rate = 100 * successful / registered,
    failure_rate = 100 * failed / registered
  ) %>%
  select(Institution, registered, successful, failed, success_rate) %>%
  arrange(desc(registered))

# Show top 30 institutions
detailed_stats %>%
  head(30) %>%
  kable(
    digits = 1,
    col.names = c("Institution", "Registered", "Successful", "Failed", "Success Rate (%)"),
    caption = "Detailed Statistics for Top 30 Institutions"
  )
```

### Distribution of Success Rates

```{r success-rate-distribution, fig.width=10, fig.height=6}
ggplot(institution_summary %>% mutate(success_rate = 100 * successful / registered), 
       aes(x = success_rate)) +
  geom_histogram(binwidth = 5, fill = "#3498db", color = "white") +
  labs(
    title = "Distribution of Success Rates Across Institutions",
    x = "Success Rate (%)",
    y = "Number of Institutions"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

## Session Info

```{r session-info}
sessionInfo()
```

## Notes

- **Registered specimens**: Specimens with any data in the `COI-5P Seq. Length` column
- **Successfully sequenced**: Specimens where `COI-5P Seq. Length` is not `0[n]` (meaning a sequence was recovered)
- **Failed/Not sequenced**: Specimens registered but with `COI-5P Seq. Length` equal to `0[n]`
- The `0[n]` notation indicates 0 base pairs with n ambiguous bases (i.e., no sequence recovered)
- Success rates vary significantly across institutions, reflecting differences in specimen quality, age, preservation methods, and sequencing protocols
