---
title: "BGE Skimming Analytics: Effect of Specimen Age on Sequencing Success"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This analysis examines the effect of specimen age on sequencing success in the BGE genome skimming project. The primary criterion for sequencing success is the number of reads aligned (`n_reads_aligned`), as older specimens might yield fewer intact reads from the target organism due to DNA degradation over time.

### Hypothesis

We expect that older specimens might yield fewer intact reads from the target organism, as:
- DNA degrades over time
- Older specimens may have relatively more contamination from organisms living on the specimen
- The sequencing coverage is still the same, but would include relatively more contaminants compared to recoverable reads from the target organism

### Methodology

For this analysis, we:
1. Aggregate data at the specimen level (group_id)
2. Only consider cases where `n_reads_aligned` is specified (not NA)
3. For each specimen, consider the assembly attempt with the highest `n_reads_aligned` value
4. Calculate specimen age from the Collection Date
5. Analyze the relationship between specimen age and sequencing success

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(lubridate)
```

## Read Data Files

### Main Analytics Data

Read the compressed, frictionless data table:

```{r read-main-data}
# Read the main analytics data
analytics_data <- read_tsv("../data/bge-skimming-analytics.tsv.gz")

cat("Analytics data dimensions:", nrow(analytics_data), "rows,", 
    ncol(analytics_data), "columns\n")
cat("Rows with n_reads_aligned:", sum(!is.na(analytics_data$n_reads_aligned)), "\n")
```

### BOLD Metadata Files

Read the BOLD metadata files needed for this analysis:

```{r read-bold-metadata}
# Read BOLD metadata files
lab_data <- read_tsv("../metadata/bold/lab.tsv")
collection_data <- read_tsv("../metadata/bold/collection_data.tsv")

cat("Lab data:", nrow(lab_data), "rows\n")
cat("Collection data:", nrow(collection_data), "rows\n")
```

## Join Data

Join the analytics data with BOLD metadata to get collection dates:

```{r join-data}
# Join analytics data with lab data to get Sample ID
enriched_data <- analytics_data %>%
  left_join(
    lab_data %>% select(`Process ID`, `Sample ID`),
    by = c("group_id" = "Process ID")
  )

# Join with collection data to get collection date
enriched_data <- enriched_data %>%
  left_join(
    collection_data %>% select(`Sample ID`, `Collection Date`),
    by = "Sample ID"
  )

cat("Enriched data dimensions:", nrow(enriched_data), "rows\n")
cat("Rows with Collection Date:", sum(!is.na(enriched_data$`Collection Date`)), "\n")
cat("Rows with both Collection Date and n_reads_aligned:", 
    sum(!is.na(enriched_data$`Collection Date`) & !is.na(enriched_data$n_reads_aligned)), "\n")
```

## Data Preparation

### Calculate Specimen Age

Parse collection dates and calculate specimen age:

```{r calculate-age}
# Current year for age calculation
current_year <- year(Sys.Date())

# Parse collection dates (format: dd-Mmm-yyyy) and calculate age
enriched_data <- enriched_data %>%
  mutate(
    # Extract year from collection date
    Collection_Year = as.integer(sub(".*-(\\d{4})$", "\\1", `Collection Date`)),
    # Calculate specimen age in years
    Specimen_Age = current_year - Collection_Year
  )

cat("Specimens with valid collection year:", sum(!is.na(enriched_data$Collection_Year)), "\n")
cat("Collection year range:", 
    min(enriched_data$Collection_Year, na.rm = TRUE), "-",
    max(enriched_data$Collection_Year, na.rm = TRUE), "\n")
cat("Specimen age range:", 
    min(enriched_data$Specimen_Age, na.rm = TRUE), "-",
    max(enriched_data$Specimen_Age, na.rm = TRUE), "years\n")
```

### Aggregate at Specimen Level

For each specimen (group_id), select the assembly attempt with the highest n_reads_aligned:

```{r aggregate-specimens}
# Filter to only specimens with n_reads_aligned specified
specimen_data <- enriched_data %>%
  filter(!is.na(n_reads_aligned)) %>%
  filter(!is.na(Specimen_Age))

cat("Specimens with n_reads_aligned and age:", nrow(specimen_data), "rows\n")

# For each specimen, select the assembly attempt with highest n_reads_aligned
specimen_best <- specimen_data %>%
  group_by(group_id) %>%
  slice_max(n_reads_aligned, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("\nAggregated to specimen level:", nrow(specimen_best), "specimens\n")
cat("Unique specimens:", n_distinct(specimen_best$group_id), "\n")
```

## Summary Statistics

### Overall Statistics

```{r overall-stats}
cat("\n=== Overall Statistics ===\n\n")
cat("Total specimens analyzed:", nrow(specimen_best), "\n")
cat("\nSequencing Success (n_reads_aligned):\n")
cat("  Mean:", sprintf("%.1f", mean(specimen_best$n_reads_aligned, na.rm = TRUE)), "reads\n")
cat("  Median:", sprintf("%.1f", median(specimen_best$n_reads_aligned, na.rm = TRUE)), "reads\n")
cat("  Min:", min(specimen_best$n_reads_aligned, na.rm = TRUE), "reads\n")
cat("  Max:", max(specimen_best$n_reads_aligned, na.rm = TRUE), "reads\n")
cat("  SD:", sprintf("%.1f", sd(specimen_best$n_reads_aligned, na.rm = TRUE)), "reads\n")

cat("\nSpecimen Age:\n")
cat("  Mean:", sprintf("%.1f", mean(specimen_best$Specimen_Age, na.rm = TRUE)), "years\n")
cat("  Median:", sprintf("%.1f", median(specimen_best$Specimen_Age, na.rm = TRUE)), "years\n")
cat("  Min:", min(specimen_best$Specimen_Age, na.rm = TRUE), "years\n")
cat("  Max:", max(specimen_best$Specimen_Age, na.rm = TRUE), "years\n")
cat("  SD:", sprintf("%.1f", sd(specimen_best$Specimen_Age, na.rm = TRUE)), "years\n")
```

### Statistics by Age Groups

```{r age-group-stats}
# Create age groups for easier analysis
specimen_best <- specimen_best %>%
  mutate(
    Age_Group = cut(Specimen_Age, 
                   breaks = c(-Inf, 5, 10, 20, 30, 50, Inf),
                   labels = c("0-5 years", "6-10 years", "11-20 years", 
                             "21-30 years", "31-50 years", ">50 years"),
                   right = TRUE)
  )

# Summary by age group
age_group_summary <- specimen_best %>%
  group_by(Age_Group) %>%
  summarise(
    n_specimens = n(),
    mean_reads = mean(n_reads_aligned, na.rm = TRUE),
    median_reads = median(n_reads_aligned, na.rm = TRUE),
    sd_reads = sd(n_reads_aligned, na.rm = TRUE),
    min_reads = min(n_reads_aligned, na.rm = TRUE),
    max_reads = max(n_reads_aligned, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n=== Statistics by Age Group ===\n\n")
kable(age_group_summary, 
      digits = 1,
      col.names = c("Age Group", "N Specimens", "Mean Reads", "Median Reads", 
                    "SD Reads", "Min Reads", "Max Reads"),
      caption = "Sequencing Success by Specimen Age Group")
```

## Visualizations

### Scatter Plot: Specimen Age vs. Reads Aligned

```{r scatter-age-reads, fig.width=12, fig.height=8}
# Create scatter plot with trend line
ggplot(specimen_best, aes(x = Specimen_Age, y = n_reads_aligned)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE, linewidth = 1.5) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Effect of Specimen Age on Sequencing Success",
    subtitle = "Each point represents the best assembly attempt for one specimen",
    x = "Specimen Age (years)",
    y = "Number of Reads Aligned"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

### Box Plot: Reads Aligned by Age Group

```{r boxplot-age-groups, fig.width=12, fig.height=8}
ggplot(specimen_best, aes(x = Age_Group, y = n_reads_aligned, fill = Age_Group)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(
    title = "Distribution of Sequencing Success by Specimen Age Group",
    x = "Specimen Age Group",
    y = "Number of Reads Aligned"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

### Bar Chart: Mean Reads Aligned by Age Group

```{r barplot-mean-reads, fig.width=12, fig.height=8}
ggplot(age_group_summary, aes(x = Age_Group, y = mean_reads, fill = Age_Group)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_reads - sd_reads, ymax = mean_reads + sd_reads),
                width = 0.2, alpha = 0.7) +
  geom_text(aes(label = sprintf("%.0f", mean_reads)), 
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(
    title = "Mean Sequencing Success by Specimen Age Group",
    subtitle = "Error bars show ± 1 standard deviation",
    x = "Specimen Age Group",
    y = "Mean Number of Reads Aligned"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

### Histogram: Distribution of Specimen Ages

```{r histogram-ages, fig.width=12, fig.height=6}
ggplot(specimen_best, aes(x = Specimen_Age)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Specimen Ages",
    x = "Specimen Age (years)",
    y = "Number of Specimens"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

### Hexbin Plot: Density Visualization

```{r hexbin-age-reads, fig.width=12, fig.height=8}
ggplot(specimen_best, aes(x = Specimen_Age, y = n_reads_aligned)) +
  geom_hex(bins = 30) +
  geom_smooth(method = "loess", color = "red", se = FALSE, linewidth = 1.5) +
  scale_fill_viridis_c(option = "plasma", trans = "log10") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Density of Specimens: Age vs. Sequencing Success",
    subtitle = "Hexagonal bins show concentration of specimens",
    x = "Specimen Age (years)",
    y = "Number of Reads Aligned",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

## Statistical Analysis

### Correlation Analysis

```{r correlation}
cat("\n=== Correlation Analysis ===\n\n")

# Pearson correlation
cor_pearson <- cor.test(specimen_best$Specimen_Age, specimen_best$n_reads_aligned, 
                       method = "pearson")
cat("Pearson Correlation:\n")
cat("  Correlation coefficient:", sprintf("%.4f", cor_pearson$estimate), "\n")
cat("  P-value:", format.pval(cor_pearson$p.value), "\n")
cat("  95% CI: [", sprintf("%.4f", cor_pearson$conf.int[1]), ", ", 
    sprintf("%.4f", cor_pearson$conf.int[2]), "]\n\n")

# Spearman correlation (robust to outliers)
cor_spearman <- cor.test(specimen_best$Specimen_Age, specimen_best$n_reads_aligned, 
                        method = "spearman", exact = FALSE)
cat("Spearman Correlation:\n")
cat("  Correlation coefficient:", sprintf("%.4f", cor_spearman$estimate), "\n")
cat("  P-value:", format.pval(cor_spearman$p.value), "\n\n")
```

### Linear Regression

```{r linear-regression}
cat("\n=== Linear Regression Analysis ===\n\n")

# Simple linear regression
lm_model <- lm(n_reads_aligned ~ Specimen_Age, data = specimen_best)
cat("Linear Model: n_reads_aligned ~ Specimen_Age\n\n")
print(summary(lm_model))

# Extract key statistics
coefficients <- coef(lm_model)
r_squared <- summary(lm_model)$r.squared
adj_r_squared <- summary(lm_model)$adj.r.squared

cat("\n=== Key Statistics ===\n")
cat("Intercept:", sprintf("%.2f", coefficients[1]), "\n")
cat("Slope:", sprintf("%.2f", coefficients[2]), "reads per year\n")
cat("R-squared:", sprintf("%.4f", r_squared), "\n")
cat("Adjusted R-squared:", sprintf("%.4f", adj_r_squared), "\n")
```

### ANOVA by Age Group

```{r anova-age-group}
cat("\n=== ANOVA: Comparing Age Groups ===\n\n")

# One-way ANOVA
anova_model <- aov(n_reads_aligned ~ Age_Group, data = specimen_best)
cat("ANOVA: n_reads_aligned ~ Age_Group\n\n")
print(summary(anova_model))

# Post-hoc test (Tukey HSD)
cat("\n=== Post-hoc Test (Tukey HSD) ===\n\n")
tukey_result <- TukeyHSD(anova_model)
print(tukey_result)
```

### Non-parametric Test

```{r kruskal-wallis}
cat("\n=== Kruskal-Wallis Test (Non-parametric) ===\n\n")

# Kruskal-Wallis test (non-parametric alternative to ANOVA)
kw_test <- kruskal.test(n_reads_aligned ~ Age_Group, data = specimen_best)
cat("Kruskal-Wallis test: n_reads_aligned ~ Age_Group\n\n")
print(kw_test)
```

## Interpretation and Key Findings

```{r key-findings}
cat("\n=== Key Findings ===\n\n")

# Calculate percentage change in mean reads from youngest to oldest group
youngest_mean <- age_group_summary$mean_reads[1]
oldest_mean <- age_group_summary$mean_reads[nrow(age_group_summary)]
percent_change <- ((oldest_mean - youngest_mean) / youngest_mean) * 100

cat("1. Effect of Specimen Age on Sequencing Success:\n")
if (coefficients[2] < 0) {
  cat("   - Negative relationship: Older specimens tend to have FEWER aligned reads\n")
  cat("   - Average decrease:", sprintf("%.2f", abs(coefficients[2])), "reads per year of age\n")
} else {
  cat("   - Positive relationship: Older specimens tend to have MORE aligned reads\n")
  cat("   - Average increase:", sprintf("%.2f", coefficients[2]), "reads per year of age\n")
}

cat("\n2. Magnitude of Effect:\n")
cat("   - Correlation coefficient:", sprintf("%.4f", cor_spearman$estimate), "\n")
cat("   - R-squared:", sprintf("%.4f", r_squared), 
    "(", sprintf("%.2f%%", r_squared * 100), "of variance explained)\n")

cat("\n3. Comparison Across Age Groups:\n")
cat("   - Youngest group (0-5 years) mean:", sprintf("%.1f", youngest_mean), "reads\n")
cat("   - Oldest group (>50 years) mean:", sprintf("%.1f", oldest_mean), "reads\n")
if (abs(percent_change) > 1) {
  cat("   - Change:", sprintf("%.1f%%", percent_change), "\n")
} else {
  cat("   - Change: Minimal difference between groups\n")
}

cat("\n4. Statistical Significance:\n")
if (cor_spearman$p.value < 0.001) {
  cat("   - Highly significant (p < 0.001)\n")
} else if (cor_spearman$p.value < 0.05) {
  cat("   - Significant (p < 0.05)\n")
} else {
  cat("   - Not statistically significant (p >= 0.05)\n")
}

cat("\n5. Data Coverage:\n")
cat("   - Total specimens analyzed:", nrow(specimen_best), "\n")
cat("   - Age range:", min(specimen_best$Specimen_Age), "-", 
    max(specimen_best$Specimen_Age), "years\n")
cat("   - Most specimens are from age group:", 
    age_group_summary$Age_Group[which.max(age_group_summary$n_specimens)], "\n")
```

## Conclusions

Based on this analysis of specimen age versus sequencing success (measured by number of aligned reads):

### Main Conclusions

1. **Relationship Strength**: The statistical analysis reveals the strength and direction of the relationship between specimen age and the number of reads aligned to the target organism.

2. **DNA Preservation**: The results help understand how well DNA from specimens of different ages is preserved and recoverable through genome skimming.

3. **Practical Implications**: These findings can inform:
   - Collection and preservation practices
   - Expectations for sequencing success based on specimen age
   - Resource allocation for sequencing older vs. newer specimens

### Limitations

- The analysis assumes collection date is a reliable proxy for specimen age
- Other factors (preservation method, storage conditions, taxonomic group) may also affect sequencing success
- The relationship may be non-linear, with different decay patterns for different age ranges

### Future Directions

- Include additional specimen-level factors (taxonomic group, institution, preservation method)
- Analyze interaction effects between age and other variables
- Compare with other sequencing success metrics (sequence quality, coverage, etc.)

## Session Info

```{r session-info}
sessionInfo()
```

## References

- [BGE Skimming Analytics README](../README.md)
- [BOLD Metadata README](../metadata/bold/README.md)
- [Merged TSV Documentation](../docs/MERGE_ALL_TSV.md)
