---
title: "Taxonomic Validation Success Analysis: Performance by Order"
author: "BGE Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This analysis examines how well different taxonomic groups perform in **taxonomic validation** during genome skimming, focusing on Linnean orders. The goal is to determine whether there is taxonomic signal in validation success rates - i.e., do some orders yield more successful taxonomic matches than others?

### Success Criteria

A specimen is considered **successful** if ANY of its assembly attempts achieve **taxonomic validation**, meaning:

- The expected taxon (`identification`) is found among the observed taxa (`obs_taxon`)
- This is checked by verifying if `identification` appears in the comma-separated `obs_taxon` list

Note: This analysis focuses purely on taxonomic validation, not structural validity (sequence length, ambiguous bases, or stop codons).

### Scope

- Analysis is aggregated at the level of Linnean **Orders**
- Only Orders with **at least 5 specimens** are included
- Only specimens whose process ID (group_id) occurs in the analytics data are analyzed (specimens in metadata/bold but not in analytics are disregarded)
- Only records with both **identification** and **obs_taxon** values are analyzed (empty attempts without taxon validation are disregarded)
- Aggregation at process ID level: when multiple attempts exist for the same specimen and some succeed while others fail, the successful attempt is considered

### Research Question

Do some taxonomic orders have higher taxonomic validation success rates than others? This could indicate:
1. **Reference library coverage effects**: Taxa with better representation in reference databases may yield denser, more accurate matches
2. **Taxonomic reconciliation issues**: Mismatches between the taxonomy in reference databases and BOLD registered taxonomy

## Load Required Libraries

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

## Read Data Files

### Main Analytics Data

```{r read-analytics}
# Read the main analytics data
analytics_data <- read_tsv("../data/bge-skimming-analytics.tsv.gz")

cat("Analytics data dimensions:", nrow(analytics_data), "rows,", 
    ncol(analytics_data), "columns\n")
cat("Unique specimens (group_id):", n_distinct(analytics_data$group_id), "\n")

# Filter to only records that have both identification and obs_taxon
# Because we did a 2-step process where we only did taxon validation for the 
# best structurally valid records, there are many assembly attempts where we 
# didn't do taxon validation at all. Disregard those empty attempts.
analytics_data_filtered <- analytics_data %>%
  filter(!is.na(identification) & identification != "" &
         !is.na(obs_taxon) & obs_taxon != "")

cat("\nAfter filtering to records with both identification and obs_taxon:\n")
cat("  Records:", nrow(analytics_data_filtered), "rows\n")
cat("  Unique specimens (group_id):", n_distinct(analytics_data_filtered$group_id), "\n")
```

### BOLD Metadata Files

```{r read-bold-metadata}
# Read BOLD metadata files needed for taxonomic information
lab_data <- read_tsv("../metadata/bold/lab.tsv")
taxonomy_data <- read_tsv("../metadata/bold/taxonomy.tsv")

cat("Lab data:", nrow(lab_data), "rows\n")
cat("Taxonomy data:", nrow(taxonomy_data), "rows\n")

# Filter BOLD metadata to only include specimens that are in our analytics data
# Only look at those specimens (group_id) that are listed within the analytics file
specimens_in_analytics <- unique(analytics_data_filtered$group_id)
lab_data_filtered <- lab_data %>%
  filter(`Process ID` %in% specimens_in_analytics)

cat("\nAfter filtering BOLD metadata to specimens in analytics:\n")
cat("  Lab data records:", nrow(lab_data_filtered), "\n")
```

## Join Data and Add Taxonomic Information

```{r join-data}
# Join analytics data with lab data to get Sample ID
enriched_data <- analytics_data_filtered %>%
  left_join(
    lab_data_filtered %>% select(`Process ID`, `Sample ID`),
    by = c("group_id" = "Process ID")
  )

# Join with taxonomy data to get taxonomic Order
enriched_data <- enriched_data %>%
  left_join(
    taxonomy_data %>% select(`Sample ID`, Phylum, Class, Order),
    by = "Sample ID"
  )

cat("Enriched data:", nrow(enriched_data), "rows\n")
cat("Rows with taxonomic Order:", sum(!is.na(enriched_data$Order)), 
    sprintf(" (%.1f%%)\n", 100 * sum(!is.na(enriched_data$Order)) / nrow(enriched_data)))
```

## Calculate Taxonomic Validation Success per Assembly Attempt

```{r calculate-success}
# Define taxonomic validation success for each assembly attempt
# Success = identification taxon appears in the obs_taxon comma-separated list
# The grepl pattern ensures exact matching: identification must be a complete 
# element in the comma-separated obs_taxon list
enriched_data <- enriched_data %>%
  mutate(
    taxonomic_valid = grepl(paste0("(^|,)", identification, "(,|$)"), obs_taxon)
  )

cat("Assembly attempts with taxonomic validation success:", 
    sum(enriched_data$taxonomic_valid, na.rm = TRUE),
    sprintf(" (%.1f%%)\n", 100 * sum(enriched_data$taxonomic_valid, na.rm = TRUE) / nrow(enriched_data)))
```

## Aggregate by Specimen (Process ID)

```{r aggregate-by-specimen}
# For each specimen (group_id), aggregate taxonomic validation results
# In cases where multiple taxon validation attempts were made for the same group_id 
# (i.e. 1step analysis), only consider one case. In cases where some attempts 
# succeeded while others failed in the 1step approach, just consider the successful attempt.
specimen_success <- enriched_data %>%
  filter(!is.na(Order)) %>%  # Only specimens with taxonomic information
  group_by(group_id, Order) %>%
  summarise(
    n_attempts = n(),
    n_successful_attempts = sum(taxonomic_valid, na.rm = TRUE),
    any_success = any(taxonomic_valid, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n=== Specimen-level Taxonomic Validation Success ===\n\n")
cat("Total specimens with Order info:", nrow(specimen_success), "\n")
cat("Specimens with at least one taxonomically validated attempt:", sum(specimen_success$any_success),
    sprintf(" (%.1f%%)\n", 100 * sum(specimen_success$any_success) / nrow(specimen_success)))
cat("Mean assembly attempts per specimen:", sprintf("%.1f\n", mean(specimen_success$n_attempts)))
```

## Filter to Orders with at Least 5 Specimens

```{r filter-orders}
# Count specimens per Order
order_counts <- specimen_success %>%
  group_by(Order) %>%
  summarise(n_specimens = n(), .groups = "drop")

# Identify Orders with at least 5 specimens
orders_to_include <- order_counts %>%
  filter(n_specimens >= 5) %>%
  pull(Order)

cat("Total Orders with taxonomic data:", nrow(order_counts), "\n")
cat("Orders with at least 5 specimens:", length(orders_to_include), "\n\n")

# Filter specimen data to only include these Orders
specimen_success_filtered <- specimen_success %>%
  filter(Order %in% orders_to_include)

cat("Specimens after filtering to Orders with >=5 specimens:", 
    nrow(specimen_success_filtered), "\n")
```

## Calculate Success Rates by Order

```{r success-by-order}
# Calculate taxonomic validation success rate for each Order
order_success <- specimen_success_filtered %>%
  group_by(Order) %>%
  summarise(
    n_specimens = n(),
    n_successful = sum(any_success),
    success_rate = n_successful / n_specimens,
    .groups = "drop"
  ) %>%
  arrange(desc(success_rate))

cat("\n=== Taxonomic Validation Success Rates by Order ===\n\n")
print(kable(order_success, digits = 3, 
            col.names = c("Order", "Specimens", "Successful", "Success Rate")))
```

## Statistical Analysis

### Overall Statistics

```{r overall-stats}
cat("\n=== Overall Statistics ===\n\n")
cat("Overall taxonomic validation success rate:", 
    sprintf("%.1f%%\n", 100 * sum(specimen_success_filtered$any_success) / nrow(specimen_success_filtered)))
cat("Range of success rates by Order:", 
    sprintf("%.1f%% to %.1f%%\n", 
            100 * min(order_success$success_rate), 
            100 * max(order_success$success_rate)))
cat("Standard deviation of success rates:", 
    sprintf("%.3f\n", sd(order_success$success_rate)))
```

### Chi-square Test for Independence

Test whether taxonomic validation success rate is independent of taxonomic Order:

```{r chi-square-test}
# Create contingency table: Order vs Success
contingency_table <- specimen_success_filtered %>%
  group_by(Order, any_success) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = any_success, values_from = n, values_fill = 0) %>%
  select(-Order) %>%
  as.matrix()

rownames(contingency_table) <- specimen_success_filtered %>%
  group_by(Order, any_success) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = any_success, values_from = n, values_fill = 0) %>%
  pull(Order)

# Perform chi-square test
chi_test <- chisq.test(contingency_table)

cat("\n=== Chi-square Test of Independence ===\n\n")
cat("Null hypothesis: Taxonomic validation success rate is independent of taxonomic Order\n\n")
cat("Chi-square statistic:", sprintf("%.2f\n", chi_test$statistic))
cat("Degrees of freedom:", chi_test$parameter, "\n")
cat("P-value:", sprintf("%.4e\n", chi_test$p.value))
cat("\nConclusion:", 
    if(chi_test$p.value < 0.001) {
      "STRONG evidence of taxonomic signal (p < 0.001)"
    } else if(chi_test$p.value < 0.05) {
      "Evidence of taxonomic signal (p < 0.05)"
    } else {
      "No significant taxonomic signal (p >= 0.05)"
    })
cat("\n")
```

### Effect Size (Cramér's V)

```{r cramers-v}
# Calculate Cramér's V as a measure of effect size
n <- sum(contingency_table)
min_dim <- min(nrow(contingency_table) - 1, ncol(contingency_table) - 1)
cramers_v <- sqrt(chi_test$statistic / (n * min_dim))

cat("\n=== Effect Size (Cramér's V) ===\n\n")
cat("Cramér's V:", sprintf("%.3f\n", cramers_v))
cat("Interpretation:", 
    if(cramers_v < 0.1) {
      "Small effect"
    } else if(cramers_v < 0.3) {
      "Medium effect"
    } else {
      "Large effect"
    })
cat("\n")
```

## Visualizations

### Success Rate by Order (Bar Plot)

```{r plot-success-rates, fig.width=12, fig.height=10}
# Order by success rate for visualization
order_success_plot <- order_success %>%
  arrange(success_rate) %>%
  mutate(Order = factor(Order, levels = Order))

ggplot(order_success_plot, aes(x = Order, y = success_rate)) +
  geom_col(aes(fill = success_rate), show.legend = TRUE) +
  geom_text(aes(label = sprintf("%.1f%%", 100 * success_rate)), 
            hjust = -0.1, size = 3) +
  geom_text(aes(label = sprintf("n=%d", n_specimens)), 
            hjust = 1.1, size = 2.5, color = "white") +
  coord_flip() +
  scale_fill_gradient(low = "#e74c3c", high = "#2ecc71", 
                      labels = scales::percent,
                      name = "Success Rate") +
  scale_y_continuous(labels = scales::percent, 
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Taxonomic Validation Success Rate by Order",
    subtitle = sprintf("Orders with ≥5 specimens (n=%d orders, %d specimens total)", 
                      nrow(order_success), sum(order_success$n_specimens)),
    x = "Taxonomic Order",
    y = "Success Rate (% of specimens with at least one taxonomically validated assembly)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 8),
    legend.position = "right"
  )
```

### Distribution of Success Rates

```{r plot-distribution, fig.width=10, fig.height=6}
# Histogram of success rates
ggplot(order_success, aes(x = success_rate)) +
  geom_histogram(binwidth = 0.05, fill = "#3498db", color = "white") +
  geom_vline(aes(xintercept = mean(success_rate)), 
             color = "#e74c3c", linetype = "dashed", size = 1) +
  annotate("text", 
           x = mean(order_success$success_rate), 
           y = Inf, 
           label = sprintf("Mean: %.1f%%", 100 * mean(order_success$success_rate)),
           vjust = 2, hjust = -0.1, color = "#e74c3c") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Distribution of Taxonomic Validation Success Rates Across Orders",
    x = "Success Rate",
    y = "Number of Orders"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )
```

### Sample Size vs Success Rate

```{r plot-sample-size, fig.width=10, fig.height=6}
ggplot(order_success, aes(x = n_specimens, y = success_rate)) +
  geom_point(aes(size = n_specimens, color = success_rate), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "#2c3e50") +
  scale_color_gradient(low = "#e74c3c", high = "#2ecc71", 
                       labels = scales::percent,
                       name = "Success Rate") +
  scale_x_log10() +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(name = "Sample Size") +
  labs(
    title = "Sample Size vs Success Rate by Order",
    subtitle = "Is success rate related to sample size?",
    x = "Number of Specimens (log scale)",
    y = "Success Rate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10)
  )
```

## Top and Bottom Performing Orders

### Top 10 Orders by Taxonomic Validation Success Rate

```{r top-orders}
cat("\n=== Top 10 Orders by Success Rate ===\n\n")
top_orders <- order_success %>%
  arrange(desc(success_rate)) %>%
  head(10)

print(kable(top_orders, digits = 3,
            col.names = c("Order", "Specimens", "Successful", "Success Rate")))
```

### Bottom 10 Orders by Taxonomic Validation Success Rate

```{r bottom-orders}
cat("\n=== Bottom 10 Orders by Success Rate ===\n\n")
bottom_orders <- order_success %>%
  arrange(success_rate) %>%
  head(10)

print(kable(bottom_orders, digits = 3,
            col.names = c("Order", "Specimens", "Successful", "Success Rate")))
```

## Summary and Conclusions

```{r summary}
cat("\n=== SUMMARY ===\n\n")
cat("1. Dataset Scope:\n")
cat("   - Total specimens analyzed:", nrow(specimen_success_filtered), "\n")
cat("   - Taxonomic Orders included:", length(orders_to_include), 
    "(with ≥5 specimens each)\n")
cat("   - Total assembly attempts:", sum(specimen_success_filtered$n_attempts), "\n\n")

cat("2. Overall Taxonomic Validation Success:\n")
cat("   - Overall specimen success rate:", 
    sprintf("%.1f%%\n", 100 * sum(specimen_success_filtered$any_success) / nrow(specimen_success_filtered)))
cat("   - Range across Orders:", 
    sprintf("%.1f%% to %.1f%%\n", 
            100 * min(order_success$success_rate), 
            100 * max(order_success$success_rate)))
cat("   - Coefficient of variation:", 
    sprintf("%.2f\n\n", sd(order_success$success_rate) / mean(order_success$success_rate)))

cat("3. Statistical Evidence of Taxonomic Signal:\n")
cat("   - Chi-square test p-value:", sprintf("%.4e\n", chi_test$p.value))
cat("   - Cramér's V (effect size):", sprintf("%.3f\n", cramers_v))
cat("   - Interpretation:", 
    if(chi_test$p.value < 0.001 && cramers_v >= 0.1) {
      "YES - Strong evidence that taxonomic Order significantly affects validation success rate"
    } else if(chi_test$p.value < 0.05) {
      "YES - Evidence that taxonomic Order affects validation success rate, but effect may be small"
    } else {
      "NO - No significant evidence that taxonomic Order affects validation success rate"
    })
cat("\n\n")

cat("4. Key Findings:\n")
cat("   - Best performing Order:", 
    order_success$Order[which.max(order_success$success_rate)],
    sprintf("(%.1f%% success, n=%d)\n", 
            100 * max(order_success$success_rate),
            order_success$n_specimens[which.max(order_success$success_rate)]))
cat("   - Worst performing Order:", 
    order_success$Order[which.min(order_success$success_rate)],
    sprintf("(%.1f%% success, n=%d)\n", 
            100 * min(order_success$success_rate),
            order_success$n_specimens[which.min(order_success$success_rate)]))

cat("\n5. Interpretation:\n")
cat("   The observed taxonomic effects could be due to:\n")
cat("   - Reference library coverage: Taxa with better representation may yield more accurate matches\n")
cat("   - Taxonomic reconciliation issues: Mismatches between reference DB and BOLD taxonomy\n")
cat("   - Biological factors: Some groups may have more conserved barcode regions\n")
```

## Session Info

```{r session-info}
sessionInfo()
```

## Notes

- **Process ID** in the analytics data corresponds to `group_id` column
- **Only specimens in the analytics file are analyzed**: Specimens that are in metadata/bold but not in the analytics file are disregarded
- **Only records with both identification and obs_taxon are analyzed**: Because we did a 2-step process where we only did taxon validation for the best structurally valid records, many assembly attempts don't have taxon validation at all. These empty attempts are disregarded.
- **Taxonomic validation success** is defined as having the `identification` taxon appear in the `obs_taxon` comma-separated list for at least one assembly attempt
- **Multiple attempts per specimen are aggregated**: In cases where multiple taxon validation attempts were made for the same group_id (i.e. 1step analysis), when some attempts succeeded while others failed, the successful attempt is considered for that specimen
- Only specimens with complete taxonomic Order information are included
- Orders with fewer than 5 specimens are excluded to avoid bias from small sample sizes
- The chi-square test assumes independence between specimens within each Order
- Effect size (Cramér's V) provides a measure of the strength of the association between Order and success
- Success rates may reflect reference library coverage, taxonomic reconciliation issues, or biological factors

